Towards a Verified Tableau Prover for a Ground Fragment of Set Theory

### 2023-02-08 paper/CADE/paper.tex
Begin with CADE paper

\documentclass[
  sigplan,
  10pt,
  %anonymous,
  %review,
  ]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

@@

% Autoref names
\renewcommand{\sectionautorefname}{Section}
\renewcommand{\subsectionautorefname}{Section}

@@

\newcommand{\fmAnd}[2]{#1 $\boldsymbol{\land}$ #2}
\newcommand{\fmOr}[2]{#1 $\boldsymbol{\lor}$ #2}
\newcommand{\fmNegSymbol}{\boldsymbol{\neg}}
\newcommand{\fmNeg}[1]{$\fmNegSymbol$ #1}
\newcommand{\fmAtom}{\textbf{A}}

@@

\title{Towards a Verified Tableau Prover for a Ground Fragment of Set Theory}
\author{Lukas Stevens}
\orcid{0000-0003-0222-6858}
\affiliation{%
  \institution{Technical University of Munich}
  \department{Department of Informatics}
  \streetaddress{Boltzmannstr. 3}
  \city{Garching}
  \postcode{85748}
  \country{Germany}
}
\email{lukas.stevens@in.tum.de}

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
\end{abstract}

@@

@@

\subsection{Linear Expansion Rules}
\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{p{0.21\textwidth}cp{0.18\textwidth} p{0.01\textwidth} p{0.21\textwidth}cp{0.18\textwidth}}
    \toprule
    \multicolumn{3}{c}{Propositional Rules} && \multicolumn{3}{c}{Rules for \lstinline!$\unionS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! &&
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\

    \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! &&
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\

    \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! &&
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\

    \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! &&
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$, s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\

    \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! &&
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\

    \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! &&
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p! &&&& \\[3ex]

    \multicolumn{3}{c}{Rules for \lstinline!$\interS$!} && \multicolumn{3}{c}{Rules for \lstinline!$\diffS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    \lstinline!s $\inS$ $\ton$ $\interS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$, s $\inS$ $\ttw$! &&
    \lstinline!s $\inS$ $\ton$ $\diffS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$, s $\notinS$ $\ttw$! \\

    \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$! && 
    \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$! \\

    \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$! &&
    \lstinline!s $\inS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$! \\

    \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$, s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ttw$! && 
    \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$, s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\

    \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$, s $\inS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$! && 
    \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$! \\

    \lstinline!s $\inS$ $\ton$, s $\inS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\interS$ $\ttw$! &&
    \lstinline!s $\inS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\diffS$ $\ttw$! \\[3ex]

    \multicolumn{3}{c}{Rules for \lstinline!Single!} && \multicolumn{3}{c}{Rules for \lstinline!$\eqS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! &&
    \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\

    \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! &&
    \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\

    \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! &&
    \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of \lstinline!l!, i.e.\ the set terms that occur directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
The perhaps more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$!
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\begin{lstlisting}
  t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b).
\end{lstlisting}
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.28\textwidth} >{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! & \lstinline!p $\in$ set b $\lor$ @\fmNeg{p}@ $\in$ set b! \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!@\fmNeg{p}@! & \lpipe & \lstinline!p! \\
    \end{tabular}
         & \lstinline!@\fmNeg{(\fmAnd{p}{q})}@ $\in$ set b!
         & \lstinline!@\fmNeg{p}@ $\in$ set b $\lor$ p $\in$ set b! \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ton$! & \lpipe & \lstinline!s $\notinS$ $\ton$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$ $\unionS$ $\ttw$) $\in$ set  b! \\
           \lstinline!$\ton$ $\unionS$ $\ttw$ $\in$ subterms (last b)! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ton$) $\in$ set b!
         \end{tabular}
         \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$) $\in$ set  b! \\
           \lstinline!($\ton$ $\interS$ $\ttw$) $\in$ subterms (last b)! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
        \end{tabular}
        \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$) $\in$ set  b! \\
           \lstinline!($\ton$ $\diffS$ $\ttw$) $\in$ subterms (last b)! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
         \end{tabular} \\
         \addlinespace[0.6em]

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!Var x $\inS$ $\ton$! \lstinline!Var x $\notinS$ $\ttw$! & \multirow{2}{*}{\rule[-0.4ex]{0.6pt}{4.8ex}} & \lstinline!Var x $\notinS$ $\ton$! \lstinline!Var x $\inS$ $\ttw$! \\
    \end{tabular}
         & 
         \begin{tabular}{c}
           \lstinline!($\ton$ $\neqS$ $\ttw$) $\in$ set b! \\
           \lstinline!$\ton$ $\in$ subterms (last b)! \\
           \lstinline!$\ttw$ $\in$ subterms (last b)! \\
           \lstinline!x $\notin$ vars b! \\
         \end{tabular}
         &
         \begin{tabular}{r@{\hspace{0.5em}}l}

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b $\land$! \\
                                   & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b $\land$! \\
                                   & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands}}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \textit{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \textit{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

@@

function realise :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` pwits b $\Longrightarrow$ realise x = vset {I x}
| x $\in$ subterms' b
  $\Longrightarrow$ realise t =
  &\phantom{$\Longrightarrow$ }&  vset {realise ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realise x = 0
\end{lstlisting}

@@

\begin{acks}
  The author thanks Kevin Kappelmann for his comments on a draft version of the paper.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{sources}

>>>>

\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}

\usepackage{graphicx}

@@

% citeauthor
\usepackage[numbers]{natbib}
\bibliographystyle{splncs04nat}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{booktabs}
\usepackage{multirow}

@@

% Autoref
\usepackage[hidelinks]{hyperref}
\renewcommand{\sectionautorefname}{Section}
\renewcommand{\subsectionautorefname}{Section}

@@

\newcommand{\fmAnd}[2]{#1 $\boldsymbol{\land}$ #2}
\newcommand{\fmOr}[2]{#1 $\boldsymbol{\lor}$ #2}
\newcommand{\fmNegSymbol}{\boldsymbol{\neg}}
\newcommand{\fmNeg}[1]{$\fmNegSymbol\:$#1}
\newcommand{\fmAtom}{\textbf{A}}

@@

\title{Towards a Verified Tableau Prover for a Ground Fragment of Set Theory}
\author{Lukas Stevens\orcidID{0000-0003-0222-6858}}
\institute{Technical University of Munich, Boltzmannstr.\ 3, 85748 Garching, Germany
  \email{lukas.stevens@in.tum.de}}

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  \keywords{First keyword \and Second keyword}
\end{abstract}

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.48\textwidth}m{0.02\textwidth}>{\centering\arraybackslash}p{0.48\textwidth}}
    \toprule
    Propositional Rules & & Rules for \lstinline!$\unionS$! \\
    \cmidrule{1-1}\cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$!, \newline \lstinline!s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & ~\newline $\Longrightarrow$ & ~\newline \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!$\interS$! && Rules for \lstinline!$\diffS$! \\
    \cmidrule{1-1}\cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!s $\inS$ $\ton$ $\interS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$!, \newline \lstinline!s $\inS$ $\ttw$! \\
      \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$! \\ 
      \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$! \\
      \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$!, \newline \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ttw$! \\ 
      \lstinline!s $\notinS$ $\ton$ $\interS$ $\ttw$! \newline \lstinline!s $\inS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$! \\ 
      \lstinline!s $\inS$ $\ton$, s $\inS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\interS$ $\ttw$! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!s $\inS$ $\ton$ $\diffS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$!, \newline \lstinline!s $\notinS$ $\ttw$! \\
      \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$! \\
      \lstinline!s $\inS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$! \\
      \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$!, \newline \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
      \lstinline!s $\notinS$ $\ton$ $\diffS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$! \\
      \lstinline!s $\inS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\diffS$ $\ttw$! \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule{1-1} \cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of \lstinline!l!, i.e.\ the set terms that occur directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
The perhaps more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$!
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\begin{lstlisting}
  t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b).
\end{lstlisting}
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!@\fmNeg{p}@! & \lpipe & \lstinline!p! \\
    \end{tabular}
         & \lstinline!@\fmNeg{(\fmAnd{p}{q})}@ $\in$ set b! &
          \begin{tabular}{rl}
            & \lstinline!p $\in$ set b $\lor$! \\
            $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
          \end{tabular} \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ton$! & \lpipe & \lstinline!s $\notinS$ $\ton$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$ $\unionS$ $\ttw$) $\in$ set  b! \\
           \lstinline!$\ton$ $\unionS$ $\ttw$ $\in$ subterms $\phi$! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ton$) $\in$ set b!
         \end{tabular}
         \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$) $\in$ set  b! \\
           \lstinline!($\ton$ $\interS$ $\ttw$) $\in$ subterms $\phi$! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
        \end{tabular}
        \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$) $\in$ set  b! \\
           \lstinline!($\ton$ $\diffS$ $\ttw$) $\in$ subterms $\phi$! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
         \end{tabular} \\
         \addlinespace[0.6em]

         \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.165\textwidth}c>{\raggedright\arraybackslash}p{0.165\textwidth}@{}}
      \\
      \midrule
      \lstinline!Var x $\inS$ $\ton$! \lstinline!Var x $\notinS$ $\ttw$! & \multirow{2}{*}{\rule[-0.4ex]{0.6pt}{4.8ex}} & \lstinline!Var x $\notinS$ $\ton$! \lstinline!Var x $\inS$ $\ttw$! \\
    \end{tabular}
         & 
         \begin{tabular}{c}
           \lstinline!($\ton$ $\neqS$ $\ttw$) $\in$ set b! \\
           \lstinline!$\ton$ $\in$ subterms $\phi$! \\
           \lstinline!$\ttw$ $\in$ subterms $\phi$! \\
           \lstinline!x $\notin$ vars b! \\
         \end{tabular}
         &
         \begin{tabular}{r@{\hspace{0.5em}}l}

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here.}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \textit{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \textit{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

@@

function realise :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` pwits b $\Longrightarrow$ realise x = vset {I x}
| x $\in$ subterms' b
  $\Longrightarrow$ realise t = vset {realise ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realise x = 0
\end{lstlisting}

@@

\subsubsection{Acknowledgements}
The author thanks Kevin Kappelmann for his comments on a draft version of the paper.

\bibliography{sources}
